Human Factors Influencing Visual Statistical Inference

### 2013-06-15 sociology_paper.tex
demographics

\subsection{Demographic Factors} Proportions of correct responses and the logarithm of average time taken to evaluate  each lineup are computed and their distributions are shown as boxplots in Figure \ref{fig:demographic_effect} for different demographic factor levels. Averages of these distributions are represented by dots inside the boxplots. The youngest age group (18-25) took least average time to evaluate a lineup. People from India took more time than others. People who have only a high school degree took less time and this is related to what we have seen for young age group. The time taken by male and female participants are almost similar.

@@

\begin{figure}[htbp] 
   \centering
   \includegraphics[width=4in]{demographic_effect.pdf} 
   \caption{Boxplots of average time taken in log scale and proportion correct responses of all the lineups plotted for each demographic factor variables. The dots inside the box represents mean. Demographic factors appear to be statistically significant for explaining time taken. Effect of graduate degree and India in proportion correct is much different but is actually the India effect since, Most of the graduate degree holders are coming from India.}
   \label{fig:demographic_effect}
\end{figure}

The distribution of time taken to evaluate a lineup is positively skewed. But in log scale it appears to be symmetric for all the demographic factor variables as we see in Figure \ref{fig:demographic_effect}. Mean and medians are very similar as well. This allowed us to fit linear mixed model to the log time taken with normal error structure.

We observed some differences in median proportion of correct responses for different levels of the demographic factors as in Figure \ref{fig:demographic_effect}. But the variability of proportion correct responses are huge as per design of the experiments since we have some super easy and some extremely difficult lineups. There were lineups which were not expected to be evaluated correctly since some of the null plots in those lineups were showing more signal than the actual plot itself. In those scenarios,  one may not reject the null hypothesis. We call these difficult lineup. Some of the lineups were so easy that data plot is detectable in a first glance. These were expected to be evaluated correctly 100\% of the time. This is why we see the boxes range from 0 to 1 in most of the factor levels. Unlike median, less differences are observed in the average proportion of correct responses for different factor levels.

Model \eqref{eqn:demographic_response} is fitted to the data with fixed effect factors such as age, country, education and gender. To estimate the overall factor main effect, a reduced model is fitted removing that factor from Model \eqref{eqn:demographic_response}. Analysis of variance (ANOVA)  results shown in Table~\ref{tbl:anova_factor} suggest that all the factor variables are significantly different in describing the the probability of correct response except gender. Gender does not seem to show any difference in performance. The similar ANOVA results are obtained by fitting Model \eqref{eqn:demographic_time} to the data of log time taken and are shown in Table~\ref{tbl:anova_factor}. All the factor variables are significantly different for log time taken including gender.

@@

To further illustrate how the individual factor levels differ from each other the detailed results of Models \eqref{eqn:demographic_response} and \eqref{eqn:demographic_time} are shown in Table \ref{tbl:model_result_demographics}. For parameter estimation the first factor levels shown in the Table are are used as the base line. The effect of all the demographic factors on log time taken to evaluate a lineup appears to be significant. But for probability of correct responses we observe that not all the factor levels are significant.

\begin{table}[htbp]
\centering
\caption{Parameter estimates of Models \eqref{eqn:demographic_time} and \eqref{eqn:demographic_response} fitted for avarage log time taken and probability of correct lineup evaluations respectively. For time taken all the demographic factors appear to be significant. For probability of correct response age group 36-40, rest of the world and graduate degree appears to be significantly different. Gender does not show any difference in performance.}
\scalebox{0.9}{
\begin{tabular}{rlrrrrrrrrrr}
  \hline
Demographic&& \multicolumn{4}{c} {Model \eqref{eqn:demographic_time} Log Time} & &\multicolumn{4}{c} {Model \eqref{eqn:demographic_response} Proportion Correct} \\

@@

Age group 36-40 is significantly different than the base line age group 18-25. The rest of the age levels appears to be similar in explaining the probability of correct response. For country, the rest of the world is different from USA but India appears to be similar to USA. We see from Table \ref{tbl:demographics} that most of the participants (about 90\%) are from India and USA. The rest of the 10\% data are from 76 different countries. This diversity in the rest of the world may make it different from India and USA.

The graduate degree holders are significantly different as well. The positive parameter estimate indicates that they perform better and the probability of correct response is higher than other education levels. As expected, there is no significant difference between Male and female performances.

Notice that lineup specific variance estimate is 5.259 with a standard error of 2.293 which is much higher than the other significant parameter estimates in Model \eqref{eqn:demographic_response}. This indicates that the major and most important factor affecting the probability of correct response is the lineup difficulty. This is also evident from the large variability in the proportion correct in Figure~\ref{fig:demographic_effect}. Because of this huge impact of lineup specific variance, the actual effect of other factor variables on the probability of correct response is in fact very small even though they appear to be significant. We illustrate this with the following example of the significant effect of graduate degree.

Effect of graduate degree is estimated as 0.182 which appears to be highly significant. This causes an increase of probability of correct response for graduate degree over high school by 
$$ \frac {\exp(-0.683+0.182)}{1+\exp(-0.683+0.182)}-  \frac {\exp(-0.683)}{1+\exp(-0.683)} = 0.377- 0.336 =0.042.$$  If we move away one standard deviation of lineup variation, in other words if we add 2.293 to teach of the exponent term in above equation, this increase reduces to 0.024. For two standard deviations it further reduces to 0.003 which is practically very small. The similar calculation shows the same negligible impact of age level 36-40 and country on the probability of correct response. Thus even though some of the factor levels are significant, their practical impact is negligible because of high variability of lineup difficulties.

>>>>

\subsection{Demographic Factors} Proportions of correct responses and the logarithm of average time taken to evaluate  each lineup are computed and their distributions are shown as boxplots in Figure \ref{fig:demographic_effect} for different demographic factor levels. Averages of these distributions are represented by dots inside the boxplots. The youngest age group (18-25) took least average time to evaluate a lineup. People from India took more time than others. People who have only a high school degree took less time and this is related to what we have seen for young age group. The time taken by male and female participants are almost the same.

@@

\begin{figure}[htbp] 
   \centering
   \includegraphics[width=4in]{demographic_effect.pdf} 
   \caption{Boxplots of average time taken in log scale and proportion correct responses of all the lineups plotted for each demographic factor variables. The dots inside the boxes represent means. Demographic factors appear to be statistically significant for explaining time taken. Effect of graduate degree and India in proportion correct is much different but is actually the India effect since, Most of the graduate degree holders are coming from India.}
   \label{fig:demographic_effect}
\end{figure}

The distribution of time taken to evaluate a lineup is positively skewed. But in log scale it appears to be symmetric for all the demographic factor variables as we see in Figure \ref{fig:demographic_effect}. Mean and medians are very similar as well. This allows us to fit linear mixed model to the log time taken with normal error structure.

We observed some differences in median proportion of correct responses for different levels of the demographic factors as in Figure \ref{fig:demographic_effect}. But the variability of proportion correct responses are huge as per design of the experiments since we have some very easy and some extremely difficult lineups. Some of the lineups  were not expected to be evaluated correctly as some of the null plots in those lineups were showing more signal than the actual plot itself. In those scenarios,  one may not reject the null hypothesis. We call these difficult lineup. Some of the lineups were so easy that data plot is detectable at a first glance. These were expected to be evaluated correctly 100\% of the time. This is why we see the boxes range from 0 to 1 in most of the factor levels. Unlike median, fewer differences are observed in the average proportion of correct responses for different factor levels.

Model \eqref{eqn:demographic_response} is fitted to the data with fixed effect factors such as age, country, education and gender. To estimate the overall factor main effect, a reduced model is fitted removing that factor from Model \eqref{eqn:demographic_response}. Analysis of variance (ANOVA)  results shown in Table~\ref{tbl:anova_factor} suggest that all the factor variables are significantly different in describing the the probability of correct response except gender. Gender does not show any significant difference in performance.  Similarly, ANOVA results are obtained by fitting Model \eqref{eqn:demographic_time} to the data of log time taken and are shown in Table~\ref{tbl:anova_factor}. All the factor variables are significantly different for log time taken including gender.

@@

To further illustrate how the individual factor levels differ from each other  detailed results of Models \eqref{eqn:demographic_response} and \eqref{eqn:demographic_time} are shown in Table \ref{tbl:model_result_demographics}. For parameter estimation the first factor levels shown in the Table are  used as the base line. Demographic factors have significant effects on log time taken to evaluate a lineup on all their levels. For the probability of correct responses not all the factor levels are significant.

\begin{table}[htbp]
\centering
\caption{Parameter estimates of Models \eqref{eqn:demographic_time} and \eqref{eqn:demographic_response} fitted for average log time taken and probability of correct lineup evaluations respectively. For time taken all the demographic factors are significant. For probability of correct response age group 36-40, rest of the world and graduate degree are significantly different. For gender we do not see any difference in performance.}
\scalebox{0.9}{
\begin{tabular}{rlrrrrrrrrrr}
  \hline
Demographic&& \multicolumn{4}{c} {Model \eqref{eqn:demographic_time} Log Time} & &\multicolumn{4}{c} {Model \eqref{eqn:demographic_response} Proportion Correct} \\

@@

Age group 36-40 is significantly different than the base line age group of 18-25 year olds. The other age levels appears to be similar in explaining the probability of correct response. For country, the rest of the world is different from USA but India appears to be similar to USA. We see from Table \ref{tbl:demographics} that most of the participants (about 90\%) are from India and USA. The rest of the 10\% data are from 76 different countries. This diversity in the rest of the world may make it different from India and USA.

The graduate degree holders are significantly different as well. The positive parameter estimate indicates that they perform better and the probability of correct response is higher than other education levels. As expected, there is no significant difference between male and female performances.

Notice that lineup specific variance estimate is 5.259 with a standard error of 2.293 which is much higher than the other significant parameter estimates in Model \eqref{eqn:demographic_response}. This indicates that the major and most important factor affecting the probability of correct response is  lineup difficulty. This is also evident from the large variability in the proportion correct in Figure~\ref{fig:demographic_effect}. Because of this huge impact of lineup specific variance, the practical impact of other factor variables on the probability of correct response is in fact very small. We illustrate this with the following example of the significant effect of graduate degree.

Effect of graduate degree is estimated as 0.182 which is highly significant. For an 18-25 year old female in the United States, the expected difference in the probability of correct response, if she has a graduate degree as compared to a high school degree, is, for a lineup with random effect of zero:
$$ \frac {\exp(-0.683+0.182)}{1+\exp(-0.683+0.182)}-  \frac {\exp(-0.683)}{1+\exp(-0.683)} = 0.377- 0.336 =0.042.$$  
For a lineup with a predicted random effect of one standard deviation, i.e. we move away one standard deviation of lineup variation by adding 2.293 to each of the exponent terms in the equation, this difference reduces to 0.024. For two standard deviations it further reduces to 0.003 which is practically very small. 
A similar calculation shows the same negligible impact of age level 36-40 (0.0533 at most)  and country (0.0424 at most) on the probability of correct response. Thus even though some of the factor levels are significant, their practical impact is negligible because of high variability of lineup difficulties.
